\pkg{digiRhythm}: an \proglang{R} package for rhythmic behavior assessment in livestock

### 2023-05-15 jstatsoft/article.Rnw
Added the article updated by Marianne.

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{\pkg{digiRhythm}: an \proglang{R} package for rhythmicity assessment using the degree of functional coupling}
\Plaintitle{digiRhythm: an R package for rhythmicity assessment using the degree of functional coupling}
\Shorttitle{digiRhythm: an R package for rhythmicity assessment using the degree of functional coupling}

%% - \Abstract{} almost as usual # comr: changed abstract throughout.
\Abstract{
Indicators that allow to reliably quantify the welfare of animals are scarce. One method that has been proposed to serve this purpose is measuring the organisms rhythmicity from activities acquired with sensors (e.g., accelerometers). The Degree of Functional Coupling (DFC) is one of these rhythmicity indicators. Computing the DFC involves several computational steps, such as computing a frequency spectrum and determining the statistical significance of its peaks. In this paper, we present the \proglang{R} package \pkg{digiRhythm}, which provides functions that allow calculating rhythmicity from activity data. Further, additional functions, such as creating actograms, computing and visualizing average activities, and the calculation of a diurnality index are also integrated in this package. The package is customized for integration in the research workflow, from raw data to creating publication-ready content and allows the simple application of these sophisticated methods.
}

@@

\section[Introduction: The Degree of Functional Coupling as a welfare indicator]{Introduction: Degree of functional coupling and animal welfare} \label{sec:intro}

The Degree of functional coupling (DFC) is a metric used to describe the extent to which a measured activity is significantly cyclic to 24 hours and thus harmonically synchronized with the periodicity of the environment. The activity can be, e.g., the number of steps during a time interval or an acceleration measured with an accelerometer mounted on the animal. The DFC was first introduced by \cite{sinz1976systemanalyse} to gain insight into the circadian rhythm of sheep and used by \cite{scheibe1999comparative} to zoo animals and wildlife by applying a sequence of mathematical operations on continuous signals collected via animal-borne sensors. With the increasing availability of sensor systems to collect behavioral data continuously, more studies using the DFC were undertaken \citep{sarout2018assessment, fuchs2022detecting}. Throughout multiple studies (some refs), it has been proven that the DFC can be used as a meaningful indicator to assess an animal's welfare, thus the importance of having an open source package specifically for this type of computation. Therefore, an \proglang{R} package was developed and published with name \pkg{digiRhythm} \citep{dgm}.

Two design considerations were followed while developing the package to avoid errors and make the library's outputs easily integrated with third-party \proglang{R} packages and other analysis workflows. The first consideration concerns the input data, whereas a specific tabular data structure is required to avoid function misuse and error propagation. The second consideration deals with the user-friendliness of the data interoperability, where the functions returns are designed to be easily exploited and integrated with other analysis workflows.

This paper is organized as follows. In section \ref{sec:dfc}, the mathematical details of computing the DFC are explained. A deep dive into the statistical significance of frequency peaks needed for the DFC computation is also highlighted. In section \ref{sec:usage}, a detailed overview on all the \pkg{digiRhythm} package functionalities is provided. Finally, conclusions are drawn in section \ref{sec:summary}.

@@

\section{Discussion} \label{sec:summary}
The \pkg{digiRhythm} package can make a significant contribution to the animal welfare assessment as it makes the analysis of animal rhythmicity more accessible. The tools is easy to uses and comes in handy especially it is delivered as an \proglang{R} package. The theoretical details and intuition behind the DFC were explained in section \ref{sec:dfc}. The usage of the package was explained in the section \ref{sec:usage}. Further usage examples could be found in the library's vignettes.

>>>>

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{\pkg{digiRhythm}: an \proglang{R} package for rhythmic behavior assessment in livestock}
\Plaintitle{digiRhythm: an R package for rhythmic behavior assessment in livestock}
\Shorttitle{digiRhythm: an R package for rhythmic behavior assessment in livestock}

%% - \Abstract{} almost as usual # comr: changed abstract throughout.
\Abstract{
Measuring an organisms rhythmicity can provide information on its physiological and psychological state. What makes this measure so interesting, is that it can be computed from any activity related time series data (e.g., accelerometers, gps tracks, smart watches (step counts), etc.). The Degree of Functional Coupling (DFC) is an established method to calculate rhythmicity. Computing the DFC involves several computational steps, such as computing a frequency spectrum and determining the statistical significance of its peaks. In this paper, we present the \proglang{R} package \pkg{digiRhythm}, which provides functions that allow calculating rhythmicity from times series data. Further, additional functions, such as creating actograms, computing and visualizing average activities, and the calculation of an diurnality index are also integrated in this package. The package is customized for integration in the research workflow, from raw data to creating publication-ready content and allows the simple application of these sophisticated methods.
}

@@

\section[Introduction]{Introduction: The Degree of Functional Coupling in the context of livestock production} \label{sec:intro}

%new start
The real-time monitoring of animal behaviour is becoming increasingly important in agriculture. It can provide new opportunities to enhance the efficiency and sustainability of livestock production and is used to enhance animal welfare (Wemelsfelder et al., 2001). Animal behaviour can be altered by stress, pain or disease, but monitoring these changes in welfare is still a challenge, since some behavioural definitions are unclear and not specifically linked to a unique adverse event (Theurer et al., 2013; Levitis et al., 2009). Relevant information about animal welfare can be obtained by studying individual differences in the capacity for behavioural adaptation (Cooper et al., 2005). Indeed, during the animals life, significant changes can occur due to internal and external stimuli; these alterations arouse adaptive behavioural patterns that can be monitored by analysing the rhythmicity of behaviours (Berger et al., 2003).
Due to the increasing development of ICT (Information and Communication Technology) based on small, non-invasive sensors, large amounts of animal behaviour data are regularly recorded, and can be used to calculate parameters describing biological rhythms (Theurer et al., 2013; Scheibe and Gromann, 2006; Berger et al. 2003). Scheibe et al. (1999) developed a parameter able to quantify and investigate the rhythmicity of behavioural patterns, the Degree of Functional Coupling (DFC), which is based on telemetric measurements of activity and/or feeding. The DFC provides an estimate of synchronization between an individual and its environment. It is calculated to each 24 h period within rolling seven day periods and predicts values between 0%
and 100% (Scheibe et al., 1999, Berger et al., 2003). An organism with a high DFC is highly synchronized with its environment, while lower DFC's indicate a lower synchronization. Breakdowns in the DFC response can be caused by stressors (e.g. transport, disease, etc.) that affect the time structure of behaviour (Berger et al., 2003; Scheibe et al., 1999). In fact, animals that live in stress-free conditions show activities all tuned to one zeitgeber (i. e. 24 hour periodicity), while their activity rhythm will deviate from the zeitgeber frequency if they are subjected to stressful situations (Scheibe et al., 1999; Berger et al., 2003).
Previous studies demonstrated that calves change their behaviour as illness develops (reduced feeding, less social behaviour and increased rest) as an adaptive response to cope with the disease (Owen-Ashley et al., 2006, Dantzer and Kelley, 2007; Borderas et al. 2008). A better understanding of changes in behaviour due to illness or stress of handling is valuable to improve animal welfare and productivity as well as handler safety (Grandin et al., 1989; Kilgore et al., 2005; Borderas et al., 2008; Robért et al., 2011). Bio-rhythmic analysis has the potential to provide the basis for the early detection of psychological and physiological disorders (Berger et al., 1999).
%new end

suggest to delete start:
With the progression of digital technologies livestock production has changed significantly. Sensors are now regularly mounted on animals, with the intention of improving management practices through monitoring performance, such as daily milk yields. However, existing sensor systems are also used for the early detection of physiological changes. To date, the reliability of sensor based abnormality detection still proves potential for optimization. While processes on farms, including livestock production, are increasingly automated, it is still not possible to reliably assess the welfare state of a farm animals. The Degree of functional coupling (DFC) presents a unique approach to activity evaluation. The DFC is a metric used to describe the extent to which a measured activity is significantly cyclic to 24 hours and thus harmonically synchronized with the periodicity of the environment. Such activity can be, e.g., the number of steps during a time interval or an acceleration measured with an accelerometer mounted on the animal, but could also be the visiting times of feeding troughs or watering stations. It is assumed that organisms in a good welfare state have a rhythmicity close to 1, whereas the rhythmicity of the organism drops when its physiological state is compromised. The DFC was first introduced by \cite{sinz1976systemanalyse} to gain insight into the circadian rhythm of sheep. Hereby, the authors confirmed that the DFC of sheep dropped before they died.  \cite{scheibe1999comparative} further used the  DFC to evaluate the welfare of zoo animals and wildlife by applying a sequence of mathematical operations on continuous signals collected via animal-borne sensors.
suggest to delete end:

With the increasing availability of sensor technologies to collect behavioral data continuously, further studies applying the DFC were performed \citep{sarout2018assessment, fuchs2022detecting}. These studies were able to confirm that the rhythm of animals is linked to their state of physical wellbeing. However, the application of the DFC was always limited to those with extensive knowledge of statistical programming. An open source package specifically for this type of computation opens up the opportunities for livestock research to easily apply these algorithms. Therefore, we developed the \proglang{R} package \pkg{digiRhythm} \citep{dgm}.

% Two design considerations were followed while developing the package to avoid errors and make the library's outputs easily integrated with third-party \proglang{R} packages and other analysis workflows. The first consideration concerns the input data, whereas a specific tabular data structure is required to avoid function misuse and error propagation. The second consideration deals with the user-friendliness of the data interoperability, where the functions returns are designed to be easily exploited and integrated with other analysis workflows.

This paper describes the functions of the package and is organized as follows. In section \ref{sec:dfc}, the mathematical details of computing the DFC are explained. Further, the statistical significance of frequency peaks needed for the DFC computation is explained. In section \ref{sec:usage}, a detailed overview of the \pkg{digiRhythm} package functions is provided. Finally, we indicate how this package can be useful for researchers in section \ref{sec:summary}.

@@

\section{Discussion} \label{sec:summary}
The \pkg{digiRhythm} package can make a significant contribution to the assessment of rhythmicity in animals, as it makes the analysis of animal rhythmicity more accessible to researchers. The tool is easy to use, especially because it is delivered as an \proglang{R} package. The theoretical details and intuition behind the DFC is explained in section \ref{sec:dfc}. The usage of the package is explained in the section \ref{sec:usage}. Further usage examples can be found in the library's vignettes.

The package enables the calculation of rhythmicity in, but not limited to, animals. To date, this is purely a computation of activity patterns that have been correlated with physiological welfare factors in previous studies. This indicator can be calculated from any type of regularly measured time series data. However, we have no evidence how these activity patterns are regulated by routine management practices, such as feeding or milking. Such routines could indirectly affect rhythmic activity patterns and blur physiological rhythmicity patterns. To date, the package provides the option to exclude known management time spans from the analysis. However, in order to increase the understanding of rhythmicity as a welfare indicator future studies should investigate how these management practices affect rhythmicity calculations and evaluate if this is a reasonable welfare indicator for animals depending on external management.

The required frequency of data collection is of utmost importance in animal-borne sensor technologies. Sensors are required to be small, so that they don't disturb the animal, and further expected to run without any sensor based maintenance, such as data reading or charging of the device. This in turn means that data collection frequency should be low. In the current study an extremely low data collection frequency is suggested (15 minutes) future studies should evaluate via downsampling, how low collection frequency can be set to create reliable rhythmicity calculations. Further it would be interesting to evaluate if activity data collected from different sources, such as access of feeding stations or watering troughs, compute similar rhythmicity outputs in the same organism.

Include 7 days aspect.

Additionally, it would be interesting to investigate the correlation between the DFC and rhythmicity computed on data from various computation techniques. For instance, \cite{moller2022diurnal} utilized a range of techniques, including set logic, circular cross-correlation, circular clustering, functional enrichment analyses, and least squares regression, on transcriptomic time series data from humans and baboons. Additionally, \cite{plaza2022gps} used GPS data to estimate rhythmicity. Comparing the correlations between rhythmicity measures derived from different types of data and between various techniques could be a useful research area.

The current package has been developed for farm animals, but the principles of rhythmicity apply for any organism, including humans. Therefore we highly recommend future research to evaluate the DigiRhythm package in other species. Many humans nowadays continuously collect activity data with their smart-watches or the phones. Thus, it would be interesting to evaluate the application of the DigiRhythm library in humans, offering an objective indicator of human welfare. A previous review highlights the potential of such rhythmicity computations in humans (https://journals.sagepub.com/doi/pdf/10.1177/0748730420940483).
